---
title: "Analysis of Elizabeth Warren's Presidential Campaign"
author: "Chad Peltier"
date: "3/2/2020"
output: github_document
---

On October 16th 2019, Senator Elizabeth Warren [hit 23.7% in FiveThirtyEight's weighted polling average](https://projects.fivethirtyeight.com/polls/president-primary-d/national/) of the 2020 Democratic primary. After consistently polling between five and seven percent until May 2019, she slowly gained in national polls until the [media](https://www.nytimes.com/2019/10/24/us/politics/2020-race-democrats-polls.html) began to describe Warren as the primary's co-front-runner. Her steady, five month rise had been juxtaposed against flat polling for the other major candidates, Joe Biden (who polled at 28.9% on May 1st and 27% on October 16th) and Bernie Sanders (who polled at 16.3% on May 1st and 14.6% on October 16th). While the [FiveThirtyEight primary forecast](https://projects.fivethirtyeight.com/2020-primary-forecast/) had not yet launched, it retroactively would have given Warren a nearly-equal probability of becoming the party's nominee as Joe Biden in early November 2019. In January 2020 the New York Times endorsed Warren (and Amy Klobuchar) for the nomination.

But then the [4th Democratic Debate](https://fivethirtyeight.com/live-blog/fourth-democratic-primary-debate/) happened, as nearly everyone piled on the newly-anointed (co)front-runner. While pundits generally praised Warren's performance -- [the New York Times' opinion writers gave Warren the top performance](https://www.nytimes.com/interactive/2019/10/16/opinion/debate-winners.html) on the night -- voters latched on to her rivals' criticisms, and her mid-October polls ended up as the high point of her national polling average. After disappointing finishes in Iowa, New Hampshire, and Nevada -- followed by a third-place finish in her home state of Massachusetts on Super Tuesday, Warren dropped out of the presidential race. 

What changed after that October debate? What factors prevented her consolidation of voters in late 2019 and early 2020? Almost every political news organization has posted numerous theories: Sexism [probably played](https://fivethirtyeight.com/features/did-sexism-and-fear-of-sexism-keep-warren-from-winning-the-nomination/) a [critical role](https://www.dataforprogress.org/blog/3/5/sexism-one-reason-why-warren-didnt-do-better), for starters. It likely fueled the concerns over "electability", a nebulous concept that has more to do with herd behavior than individual preferences (as my wife argued, everyone incorrectly estimating what someoneelse's preferences are) in a year in which voters have consistently argued in polling that they would prefer someone who can beat Trump over anything else. Warren was the clear favorite [magic wand](https://www.dataforprogress.org/blog/2019/7/30/ambivalent-support-why-do-primary-voters-say-theyll-vote-for-a-non-preferred-candidate) candidate -- who voters would magically, immediately make President. 

The release of the massive [Democracy Fund + UCLA Nationscape polling data](https://www.usatoday.com/in-depth/news/politics/elections/2020/02/28/nationscape-research-voters-views-on-several-policies-remain-polarized/4569383002/?link_id=10&can_id=8f8a3198bb402b42b1635b935f7d8afc&source=email-weekend-reading-march-1-2020&email_referrer=email_737723&email_subject=weekend-reading-march-1-2020) allows us to look at Warren's polling over time and in relation to numerous other demographic and cultural factors as well as policy preferences to get a better sense for why Warren's campaign hit a wall in mid-October and never recovered.

```{r message=FALSE}
library(tidycensus)
library(tidyverse)
library(tigris)
library(tidyverse)
library(sf)
library(haven)
library(viridis)
library(lubridate)
library(ggthemes)
```

First we'll read in all of the Nationscape files and combine them into a gargantuan dataset using purrr::map and haven::read_dta to handle .dta files. 

```{r warning=FALSE}
ns_list <- list.files(pattern = "*.dta")

test <- read_dta("ns20190718.dta")


# Iterate through list of files
ns_sum <- map(ns_list, read_dta) %>%
    bind_rows

write.csv(ns_sum, file = "ns_sum.csv")
```

# Warren analysis
Let's start with some basic summaries and charts about Warren's popularity.

```{r}

# Warren
ns_sum %>%
    filter((cand_favorability_warren == 3 | cand_favorability_warren == 4) & 
               cand_favorability_warren != 999 & cand_favorability_sanders != 999 &
               cand_favorability_cortez != 999 & cand_favorability_trump != 999) %>%
    summarise(sanders = mean(cand_favorability_sanders, na.rm = TRUE),
              aoc = mean(cand_favorability_cortez, na.rm = TRUE),
              trump = mean(cand_favorability_trump, na.rm = TRUE))


# Warren over time
ns_sum %>%
    filter(cand_favorability_warren != 999 & !is.na(cand_favorability_warren) & !is.na(pid3)) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+")),
           cand_favorability_warren = if_else(cand_favorability_warren == 1 | 
                                            cand_favorability_warren == 2, "positive", "negative"),
           pid3 = if_else(pid3 == 1, "Democrat", 
                          if_else(pid3 == 2, "Republican",
                                  if_else(pid3 == 3, "Independent", "Other")))) %>%
    group_by(start_date, cand_favorability_warren, pid3) %>%
    summarize(warren = n()) %>%
    mutate(warren_percent = warren/sum(warren)) %>%
    filter(cand_favorability_warren == "positive") %>%
    ggplot(aes(x = start_date, y = warren_percent)) + 
        geom_point() +
        geom_smooth() +
        facet_wrap(~ pid3) +
        scale_y_continuous(labels = scales::percent_format()) +
        ggtitle("Elizabeth Warren Favorability by Party ID")


f_ew <- ns_sum %>%
    filter(cand_favorability_warren != 999 & !is.na(cand_favorability_warren) & pid3 == 1) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+")),
           cand_favorability_warren = if_else(cand_favorability_warren == 1 | 
                                            cand_favorability_warren == 2, "positive", "negative"),
           pid3 = if_else(pid3 == 1, "Democrat", 
                          if_else(pid3 == 2, "Republican",
                                  if_else(pid3 == 3, "Independent", "Other")))) %>%
    group_by(start_date, cand_favorability_warren) %>%
    summarize(warren = n()) %>%
    #group_by(start_date) %>%
    mutate(warren_percent = warren/sum(warren)) %>%
    filter(cand_favorability_warren == "positive") %>%
    ggplot(aes(x = start_date, y = warren_percent)) + 
        geom_point() +
        geom_smooth() +
        scale_y_continuous(labels = scales::percent_format()) +
        ylim(.7, 1) +
        ggtitle("Elizabeth Warren Favorability Among Democrats")


ns_sum %>%
    filter(cand_favorability_warren != 999 & !is.na(cand_favorability_warren)) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+")),
           cand_favorability_warren = if_else(cand_favorability_warren == 1 | 
                                            cand_favorability_warren == 2, "positive", "negative")) %>%
    group_by(start_date, cand_favorability_warren) %>%
    summarize(warren = n()) %>%
    mutate(warren_percent = warren/sum(warren)) %>%
    filter(cand_favorability_warren == "positive") %>%
    ggplot(aes(x = start_date, y = warren_percent)) + 
        geom_point() +
        geom_smooth() +
        scale_y_continuous(labels = scales::percent_format()) +
        ggtitle("Elizabeth Warren Favorability")


ns_sum %>%
    filter(cand_favorability_warren != 999 & !is.na(cand_favorability_warren)) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+")),
           cand_favorability_warren = if_else(cand_favorability_warren == 1 | 
                                            cand_favorability_warren == 2, "positive", "negative"),
           gender = as.factor(if_else(gender==1, "Female", "Male"))) %>%
    group_by(start_date, gender, cand_favorability_warren) %>%
    summarize(warren = n()) %>%
    mutate(warren_percent = warren/sum(warren)) %>%
    filter(cand_favorability_warren == "positive") %>%
    ggplot(aes(x = start_date, y = warren_percent, color = gender)) + 
        geom_point() +
        geom_smooth(aes(group = gender)) +
        scale_y_continuous(labels = scales::percent_format()) +
        ggtitle("Elizabeth Warren Favorability by Gender")


ns_sum %>%
    filter(cand_favorability_warren != 999 & !is.na(cand_favorability_warren) & !is.na(pid3)) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+")),
           cand_favorability_warren = if_else(cand_favorability_warren == 1 | 
                                            cand_favorability_warren == 2, "positive", "negative"),
           gender = as.factor(if_else(gender==1, "Female", "Male")),
           pid3 = if_else(pid3 == 1, "Democrat", 
                          if_else(pid3 == 2, "Republican",
                                  if_else(pid3 == 3, "Independent", "Other")))) %>%
    group_by(start_date, gender, pid3, cand_favorability_warren) %>%
    summarize(warren = n()) %>%
    mutate(warren_percent = warren/sum(warren)) %>%
    filter(cand_favorability_warren == "positive") %>%
    ggplot(aes(x = start_date, y = warren_percent, color = gender)) + 
        geom_point() +
        geom_smooth(aes(group = gender)) +
        facet_wrap(~ pid3) +
        scale_y_continuous(labels = scales::percent_format()) +
        ggtitle("Elizabeth Warren Favorability by Gender")

## favorability vs. intent to vote
ns_warren_percent <- ns_sum %>%
    filter(cand_favorability_warren != 999 & !is.na(cand_favorability_warren) & pid3 == 1) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+")),
           cand_favorability_warren = if_else(cand_favorability_warren == 1 | 
                                            cand_favorability_warren == 2, "positive", "negative"),
           pid3 = if_else(pid3 == 1, "Democrat", 
                          if_else(pid3 == 2, "Republican",
                                  if_else(pid3 == 3, "Independent", "Other")))) %>%
    group_by(start_date, cand_favorability_warren) %>%
    summarize(warren = n()) %>%
    mutate(warren_percent = warren/sum(warren)) %>%
    filter(cand_favorability_warren == "positive") %>%
    select(start_date, warren_percent)

ns_warren_intent <- ns_sum %>% 
    filter(dem_vote_intent != 888 & dem_vote_intent != 999 & !is.na(dem_vote_intent)) %>%
     mutate(start_date = ymd(str_remove(start_date, "\\s.+"))) %>%
    select(start_date, dem_vote_intent) %>% 
    group_by(start_date, dem_vote_intent) %>% 
    summarize(warren_intent = n()) %>%
    group_by(start_date) %>%
    mutate(warren_intent_percent = warren_intent/sum(warren_intent)) %>%
    filter(dem_vote_intent == 8) %>%
    select(start_date, warren_intent_percent) %>%
    inner_join(ns_warren_percent, by = "start_date")


ns_warren_intent %>%
    pivot_longer(cols = c(warren_intent_percent, warren_percent), 
                 names_to = "name", values_to = "values") %>%
    ggplot(aes(x = start_date, y = values, color = name, group = name)) + 
        geom_point() +
        geom_smooth() +
        scale_y_continuous(labels = scales::percent_format()) +
        ggtitle("Elizabeth Warren Favorability vs. Intent to Vote Among Democrats")

ns_warren_intent %>%
    ggplot(aes(x = warren_percent, y = warren_intent_percent)) + 
        geom_point() +
        geom_smooth() +
        scale_y_continuous(labels = scales::percent_format()) +
        ggtitle("Elizabeth Warren Favorability vs. Intent to Vote Among Democrats")

## Liz vs. Bernie
ns_sum %>%
    filter(cand_favorability_sanders != 999 & !is.na(cand_favorability_sanders) & pid3 == 1 &
               cand_favorability_warren != 999 & !is.na(cand_favorability_warren)) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+")),
           cand_favorability_sanders = if_else(cand_favorability_sanders == 1 | 
                                            cand_favorability_sanders == 2, 1, 0),
           cand_favorability_warren = if_else(cand_favorability_warren == 1 | 
                                            cand_favorability_warren == 2, 1, 0),
           pid3 = if_else(pid3 == 1, "Democrat", 
                          if_else(pid3 == 2, "Republican",
                                  if_else(pid3 == 3, "Independent", "Other")))) %>%
    select(start_date, cand_favorability_sanders, cand_favorability_warren) %>%
    pivot_longer(cols = c(cand_favorability_sanders, cand_favorability_warren), 
                 names_to = "candidate", values_to = "favorability") %>%
    group_by(start_date, candidate) %>%
    summarize(fav = mean(favorability)) %>%
    ggplot(aes(x = start_date, y = fav, color = candidate)) + 
        geom_point() +
        geom_smooth() +
        scale_y_continuous(labels = scales::percent_format()) +
        ggtitle("Elizabeth Warren vs. Bernie Sanders Favorability Among Democrats") +
        ggthemes::theme_fivethirtyeight()
ggsave("liz_bernie.png", height = 9/1.2, width = 16/1.2)
```

In the plots above favorability is measured as a binary yes-no, and then converted to a percent, rather than along the 1-4 scale used in the original polling question.

* Not a surprise: voters who do not like Warren also did not like Sanders, liked Alexandria Ocasio-Cortez even less, and liked Trump.
* In terms of favorability by self-identified party, Democrats had fairly consistent support for Warren at around 60% favorability. Independents maintained roughly a 20-25% favorability, and Republicans were around 10%. 
* Warren's overall favorability (among all polled voters) seemed to rise with national polls until October when she declined to roughly 50%.
* Warren consistently had a higher favorability among female voters than male voters, overall, among Democrats, and among Independents, but male Republican voters had a slightly higher opinion of Warren than female Republicans.  
* Maybe the most critical charts are the second- and third-to-last, which show Warren's favorability compared with Democrats intent to vote for her in their primary or caucus. There was a consistently wide gulf between favorability and intent to vote among Democrats who intended to vote, with the voters expressing over 75% favorability but between 20% and 25% intent to vote. The pecentage of Democrats intending to vote for Warren was steadily rising from August until mid-October when they fell back to her level from late summer. Warren's favorability was much more steady despite a similar, brief decline. Comparing favorability and intent to vote directly (in the second-to-last chart) it is clear that there was no relationship between the two. 
* The final chart shows Warren and Sanders' favorability head-to-head. Warren had a consistent but narrow lead over Sanders from summer until mid-October, when Sanders overtook Warren.

All of this suggests that the critical moment for Warren's campaign was in mid-October. 

# Warren Map
Because the Nationscape data includes the individual's congressional district, we can also map her support using the tigris package to import shape files for the U.S. 

```{r}

## map
ga_cd <- congressional_districts(class = "sf", cb = TRUE, resolution = "20m")

states <- tigris::states()
states <- states@data 
states <- states %>%
    select(STATEFP, NAME)

ga_cd <- ga_cd %>% 
    left_join(states, by = "STATEFP")

state_abb <- state.abb %>%
    tibble() %>%
    cbind(state.name) %>%
    rename(state_abb = 1)

ga_cd <- ga_cd %>% 
    left_join(state_abb, by = c("NAME" = "state.name"))

ga_cd <- ga_cd %>%
    unite(col = congress_district, state_abb, CD115FP, sep = "")

head(ga_cd)

ns_warren <- ns_sum %>%
    filter(cand_favorability_warren != 999 & !is.na(cand_favorability_warren) & !is.na(pid3)) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+")),
           cand_favorability_warren = if_else(cand_favorability_warren == 1 | 
                                            cand_favorability_warren == 2, "positive", "negative"),
           gender = as.factor(if_else(gender==1, "Female", "Male")),
           pid3 = if_else(pid3 == 1, "Democrat", 
                          if_else(pid3 == 2, "Republican",
                                  if_else(pid3 == 3, "Independent", "Other")))) %>%
    select(cand_favorability_warren, gender, pid3, start_date, congress_district, state) %>%
    left_join(ga_cd, by = "congress_district")


ns_warren %>%
    group_by(congress_district, cand_favorability_warren) %>%
    summarize(n = n()) %>%
    group_by(congress_district) %>%
    summarise(percent_pos = sum(n[cand_favorability_warren=="positive"])/sum(n)) %>%
    left_join(ga_cd, by = "congress_district") %>%
    filter(!str_detect(congress_district, "HI") & !str_detect(congress_district, "AK")) %>%
    ggplot(aes(fill = percent_pos)) +
        geom_sf(aes(geometry = geometry)) +
        ggthemes::theme_map() +
        ggtitle("Warren Favorability by Congressional District") +
        labs(caption = "Data from Nationscape project. Favorability includes percent of polled with very favorable or somewhat favorable opinion.") +
        scale_fill_viridis(direction = -1)
ggsave("ns_warren_map2.png", width = 16/1.2, height = 9/1.2)



ns_warren %>%
    filter(pid3 == "Democrat") %>%
    group_by(congress_district, cand_favorability_warren) %>%
    summarize(n = n()) %>%
    group_by(congress_district) %>%
    summarise(percent_pos = sum(n[cand_favorability_warren=="positive"])/sum(n)) %>%
    left_join(ga_cd, by = "congress_district") %>%
    filter(!str_detect(congress_district, "HI") & !str_detect(congress_district, "AK")) %>%
    ggplot(aes(fill = percent_pos)) +
        geom_sf(aes(geometry = geometry)) +
        ggthemes::theme_map() +
        ggtitle("Warren Favorability Among Democrats by Congressional District") +
        labs(caption = "Data from Democracy Fund + UCLA Nationscape project. Favorability includes percent of Democrats polled with very favorable or somewhat favorable opinion.") +
        #scale_fill_viridis(direction = -1) +
        scale_fill_binned(type = "viridis", direction = -1)
ggsave("ns_warren_map3.png", width = 16/1.2, height = 9/1.2)


## intent to vote - edit below
ns_sum %>%
    filter(pid3 == 1) %>%
    group_by(congress_district, dem_vote_intent) %>%
    summarize(n = n()) %>%
    mutate(warren_intent_percent = n/sum(n)) %>%
    filter(dem_vote_intent == 8) %>%
    group_by(congress_district) %>%
    left_join(ga_cd, by = "congress_district") %>%
    filter(!str_detect(congress_district, "HI") & !str_detect(congress_district, "AK")) %>%
    ggplot(aes(fill = warren_intent_percent)) +
        geom_sf(aes(geometry = geometry)) +
        ggthemes::theme_map() +
        ggtitle("Warren Support Among Democrats by Congressional District") +
        labs(caption = "Data from Democracy Fund + UCLA Nationscape project. Support is defined as who the voter would support if the election were held today.") +
        #scale_fill_viridis(direction = -1) +
        scale_fill_binned(type = "viridis", direction = -1)
ggsave("ns_warren_map4.png", width = 16/1.2, height = 9/1.2)
```

The three maps above include Warren's favorability overall, her favorability among Democrats, and the percentage of voters who intended to vote for her. 

While Democrats had high opinions of Warren from a wide vareity of districts from all over the country, there were far fewer districts with more than 20% of polled voters who intended to vote for her (critical for collecting delegates during primaries). 

# Data pre-processing and a few more exploratory data plots
We can look at a few more exploratory data plots and pre-process the data for modeling whether a democratic primary voter intended to vote for Warren or not. As part of this process we'll add in [data on population density for congressional districts from City Lab](https://www.citylab.com/equity/2018/11/citylab-congressional-density-index/575749/).

```{r warning= FALSE}
## prep data

library(naniar)
library(magrittr)
library(tidymodels)

ns_select <- ns_sum %>%
    filter(pid7 < 4) %>%
    select(dem_vote_intent, interest, vote_intention, group_favorability_socialists, 
           group_favorability_labor_unions, cand_favorability_biden, cand_favorability_sanders,
           racial_attitudes_tryhard, gender_attitudes_maleboss, gender_attitudes_logical, 
           gender_attitudes_opportunity, gender_attitudes_complain, discrimination_blacks,
           discrimination_men, discrimination_women, ideo5, employment, foreign_born, in_union, 
           medicare_for_all, age, gender, household_income, education, congress_district) %>%
    mutate(vote_warren = if_else(dem_vote_intent == 8 , 1, 0),
           student = if_else(employment == 7, 1, 0),
           full_time = if_else(employment == 1, 1, 0),
           unemployed = if_else(employment == 4, 1, 0)) %>%
    select(-employment, - dem_vote_intent) %>%
    replace_with_na_all(condition = ~.x == "888") %>%
    replace_with_na_all(condition = ~.x == "999") #%>%
    #drop_na()


citylab <- read_csv("citylab.csv")
colnames(citylab) %<>% str_replace_all("\\s", "_") %>% tolower()
citylab <- citylab %>%
    mutate(cd = str_remove(cd, "-"),
           cd = if_else(cd == "AKAL", "AK00", cd))

ns_select <- ns_select %>%
    left_join(citylab, by = c("congress_district" = "cd"))

## remove variables with a lot of NAs (can decide about imputing later)
ns_select <- ns_select %>%
    select(-congress_district, -cluster, -group_favorability_labor_unions, -group_favorability_socialists,
           - medicare_for_all)
## edit this if decide to impute
ns_select2 <- ns_select %>%
    drop_na() %>%
    mutate(vote_warren = as.factor(if_else(vote_warren == 1, "yes", "no")))

skimr::skim(ns_select)

library(ggridges)
library(GGally)

ns_select2 %>%
    ggplot(aes(x = age, y = vote_warren, fill = vote_warren)) +
        geom_density_ridges()

ns_sum %>%
    filter(dem_vote_intent %in% c(1, 7, 8)) %>%
    ggplot(aes(x = age, y = as.factor(dem_vote_intent), fill = as.factor(dem_vote_intent))) +
        geom_density_ridges()


ns_select2 %>%
    select(vote_warren, age, gender, education, high_density, low_density, medium_density) %>%
    ggpairs(mapping = aes(color = vote_warren))
ggsave("vote_warren_ggpairs.png", height = 9/1.2, width = 16/1.2)



## second choice all candidates

second_choice <- tibble(c(1, 2, 3, 4, 7, 8, 9, 10, 11), 
                        c("Joe Biden", "Cory Booker", "Pete Buttigieg", "Julian Castro", 
                          "Bernie Sanders", "Elizabeth Warren", "Other", 
                          "Amy Klobuchar", "Mike Bloomberg"))

colnames(second_choice) <- c("rank_dems_2", "candidate_second_choice")

ns_sum %>%
    filter(rank_dems_2 != 888 & rank_dems_2 != 999 & !is.na(rank_dems_2)) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+"))) %>%
    left_join(second_choice) %>%
    group_by(start_date, candidate_second_choice) %>%
    summarize(second_pref = n()) %>%
    mutate(perc = second_pref / sum(second_pref)) %>%
    ggplot(aes(x = start_date, y = perc, color = candidate_second_choice, 
               group = candidate_second_choice)) + 
        geom_point() +
        geom_smooth(se = FALSE) +
        ylim(.1, .35)
        ggtitle("Second Choice Candidate Preferences") +
        ggthemes::theme_fivethirtyeight()
        ggsave("second_choice.png", height = 9/1.2, width = 16/1.2)
        
ns_sum %>%
    filter(rank_dems_1 != 888 & rank_dems_1 != 999 & !is.na(rank_dems_1) ) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+"))) %>%
    left_join(second_choice) %>%
    rename(candidate_first_choice = candidate_second_choice) %>%
    group_by(start_date, candidate_first_choice) %>%
    summarize(first_pref = n()) %>%
    mutate(perc = first_pref / sum(first_pref)) %>%
    filter(perc < .4) %>%
    ggplot(aes(x = start_date, y = perc, color = candidate_first_choice, 
               group = candidate_first_choice)) + 
        geom_point() +
        geom_smooth(se = FALSE) +
        ggtitle("First Choice Candidate Preferences") +
        scale_y_continuous(labels = scales::percent_format())
        ggsave("first_choice.png", height = 9/1.2, width = 16/1.2)        
        
        
        
ns_second_choice <- ns_sum %>%
    filter(rank_dems_2 != 888 & rank_dems_2 != 999 & !is.na(rank_dems_2)) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+"))) %>%
    left_join(second_choice) %>%
    group_by(start_date, candidate_second_choice) %>%
    summarize(second_pref = n()) %>%
    mutate(perc_second = second_pref / sum(second_pref)) %>%
    filter(candidate_second_choice == "Elizabeth Warren") %>%
    select(start_date, perc_second)

ns_first_choice <- ns_sum %>%
    filter(rank_dems_1 != 888 & rank_dems_1 != 999 & !is.na(rank_dems_1)) %>%
    mutate(start_date = ymd(str_remove(start_date, "\\s.+"))) %>%
    left_join(second_choice) %>%
    rename(candidate_first_choice = candidate_second_choice) %>%
    group_by(start_date, candidate_first_choice) %>%
    summarize(first_pref = n()) %>%
    mutate(perc_first = first_pref / sum(first_pref)) %>%
    filter(candidate_first_choice == "Elizabeth Warren") %>%
    select(start_date, perc_first)



ns_warren_intent2 <- ns_warren_intent %>%
    left_join(ns_first_choice, by = "start_date") %>%
    left_join(ns_second_choice, by = "start_date")

ns_warren_intent2 %>%
    pivot_longer(cols = 2:5, names_to = "stat", values_to = "values") %>%
    filter(stat != "warren_percent" & values < .4) %>%
    ggplot(aes(x = start_date, y = values, color = stat, group = stat)) +
        geom_point() +
        geom_smooth(se = FALSE) +
        theme_fivethirtyeight() +
        ggtitle("Voters Favoring Elizabeth Warren vs. Intending to Vote For Her") +
        scale_y_continuous(labels = scales::percent_format())
        ggsave("Warren_favor_intent.png", height = 9/1.2, width = 16/1.2)



```

* Warren rose as Democratic voters' first choice until plateauing in mid-October. Her percentage of voters who ranked Warren as their first choice did not decline as did her intent to vote. She was the first choice for the second-most Democratic primary voters, behind Sanders (Biden was third for much of the polling time period). Also notable was that the percentage of voters who listed "NA" declined rapidly after October. These trends were largely true also looking at voters' second choices. 
* The third chart is also critical to understanding Warren's campaign. She had a high percentage of first and second choices for voters, both of which rose until mid-October, then plateauing after. But the percentage of voters who intended to vote for her dramatically declined after October as opposed to just plateauing. 
* The ridgeline plot showing Biden, Sanders, and Warren support by age is interesting. Biden received fairly steady support among all ages, but with a slight increase among older voters. Sanders's support is very clearly concentrated among younger voters. Warren's support by age is bi-modal, with a peak of 30-ish supporters as well as older voters. This could explain a significant amount on its own, given that turnout is historically low for younger voters and higher for older voters.
* The ggpairs plot also shows notable differnces for Warren supporters in age, education (Warren voters tended to be higher educated), and in cities (shown by the "higher density" variable from City Lab).


# Model building
This model is intended to predict whether a Democrat intended to vote for Warren or not based on a variety of variable listed above, including: interest in the election, intention to vote, favorable opinion of Biden, favorable opinion of Sanders, indicators of racism, indicators of sexism, employment stats, whether they were born in the U.S. or abroad, union membership, age, gender, household income, education level, and population density of their congressional district. 


```{r cache=TRUE}

## Split data
set.seed(1234)

ns_split <- ns_select2 %>% # change this if decide to impute
    initial_split(prop = 0.8)

ns_train <- training(ns_split)
ns_train_cv <- vfold_cv(ns_train, strata = vote_warren)
ns_test <- testing(ns_split)

## Recipe
ns_rec <- recipe(vote_warren ~ ., data = ns_train) %>%
    step_corr(all_numeric()) %>%
    step_zv(all_numeric()) %>%
    step_normalize(all_numeric()) %>%
    #step_knnimpute(all_predictors()) %>%
    prep()

ns_rec

#ns_prep <- ns_rec %>% 
 #   prep()

ns_juiced <- ns_rec %>%
    juice()

## Model 1
library(tune)
rf_spec <- rand_forest(mode = "classification",
                       mtry = tune(),
                       trees = 1000,
                       min_n = tune()) %>%
    set_engine("ranger", importance = "impurity")

rf_grid <- tune_grid(
    ns_rec,
    model = rf_spec,
    resamples = ns_train_cv,
    control = control_resamples(save_pred = TRUE)
)


```

```{r}
rf_grid %>%
    collect_metrics() 

rf_grid %>%
    show_best("accuracy") 

rf_best <- rf_grid %>%
    select_best(metric = "accuracy")

final_model_rf <- rf_spec %>%
    update(mtry = rf_best$mtry, min_n = rf_best$min_n) %>%
    fit(vote_warren ~ ., data = ns_juiced)

## Variable Importance
library(vip)
final_model_rf %>%
    vip()


## Test data evaluation
final_model_rf %>%
    predict(new_data = bake(ns_rec, ns_test)) %>%
    mutate(truth = as.factor(ns_test$vote_warren)) %>%
    accuracy(truth, .pred_class)
```

The random forest model performed well on the test data, achieving nearly 84% accuracy for classifying a poll respondent as supporting Warren or not. 

The variable important plot is interesting. Age was the most important predictor, followed by all four population density variables and then household income. 

Age and income (and education) are not too surprising. While the variable importance plot doesn't tell us the direction of the effect on supporting Warren, my *guess* is that medium density districts are more supportive (wealthier suburbs), low and very low density districts are less supportive, and high density districts are more supportive. But this is something that could be corroborated by other data more analysis of the Nationscape data. 

This is helpful to both begin to understand the groups of voters who might have supported Warren. Perry Bacon Jr.'s [analysis of why Warren's campaign didn't win the nomination](https://fivethirtyeight.com/features/why-warren-couldnt-win/) illustrates the major point here: 

> But she just never caught on with a broad swath of voters — polls suggest that she had little support outside of white college graduates. The New York Times described Warren as the candidate who often had the support of the “grass tops” rather than the grassroots — meaning that the leaders of activist groups often really liked Warren, but it’s not clear that the lower ranks did. For example, Warren won the personal endorsement of the president of the American Federation of Teachers, but the union itself wouldn’t endorse her because many of its members were with Biden or Sanders.

Beyond the demographics of her support, the sudden decline in polled democrats supporting Warren in mid-October -- despite steady favorability ratings -- suggests that attacks on her *perceived* electability were really damaging. Enough people believed that electability would be a concern for *other people* that they chose not to support her as their top choice, despite still holding a favorable opinion of her as a candidate. Some of her support from younger progressives likely went to Bernie during that time, while some of her support from both older voters and 30s-ish moderate voters was divided between candidates like Mayor Pete, Amy Klobuchar, and Biden. 

And no analysis of Warren's (or any other female presidential candidate's) electability would be fair without recognizing the deep-rooted sexism at the heart of such concerns. As one voter implied (if I'm remembering the right podcast) on The Daily's episode, [The Field: What Happened to Elizabeth Warren](https://www.nytimes.com/2020/03/10/podcasts/the-daily/warren.html), it's not personal sexism -- it's concern that *other* people are sexist that hurt her support.  

